Algebraic File Synchronization: Adequacy and Completeness

### 2018-07-12 p2/def-filesystem.tex
Style improvements

\begin{mydef}[Filesystems, $\FS$]
We model a filesystem
using a function $\FS$ with a set of nodes (potential filesystem paths) $\setn$ as its domain,
and a set of possible contents or values $\setv$ as its codomain:
\[ \FS: \setn \rightarrow \setv. \]
We consider metadata to be part of the values in $\setv$.
In this model, $\setn$ 
serves as a namespace for the filesystem:
it contains all possible nodes, including the ones 
where the file system contains no file or directory.
\end{mydef}

@@

There is an ancestor / descendant relation defined over the nodes in $\setn$
which arranges them in a disjoint union of rooted directed trees,
and which can be derived from the partial function $\parent$ yielding
the parent of a node in $\setn$ provided it exists.
Tao et al. \cite{TSR} describe a similar filesystem model, although
they also model inodes and restrict the filesystem to just a single tree.

\begin{mydef}[Ordering on $\setn$: $\parent$, $\descendant$, $\descendantEq$, $\unrel$]
The partial function $\parent:\setn\nrightarrow\setn$
returns the parent node of $n$,
and is undefined if $n$ is the root of a tree.

@@

We write $n\unrel m$, or $n$ and $m$ are \emph{incomparable},
iff $n\not\descendantEq m$ and $n\not\ancestorEq m$;
that is, incomparable nodes are on different branches or on different trees.
% none of the path names describing their location is an initial segment of the other.
\end{mydef}

We assume that the $\parent$ function does not induce loops, and so
$\descendant$ is indeed a strict partial ordering.

@@

The combination of the set of nodes and the parent function,
$\langle \setn, \parent \rangle$, forms a \emph{skeleton}
which the filesystems populate with values from $\setv$.
As we require filesystem functions to be total,
we use a special value, $\empt\in\setv$, to indicate that the filesystem
is empty at a particular node, that is, there are no files or directories there.
We also assume that a filesystem has finitely many non-empty nodes.

\begin{mydef}[Filesystem values: $\vald$, $\valfx$, $\empt$, $\eqclass{v}$]
The set of values $\setv$ is partitioned into files, directories, and the empty value $\empt$.
As usual, we write $[v]$ for the the equivalence class of $v\in\setv$ according to this partition,
which represents its type.
File and directory values are also denoted by $\valfx$ and $\vald$, respectively,
with or without indices.
\end{mydef}

@@

Every filesystem has a so-called \emph{tree property}, which means that
if the filesystem is not empty at a node, and the node has a parent,
then there must be a directory at the parent node.
Using the notation introduced above, we can formally express this as follows.
\begin{mydef}[Tree property]
A filesystem $\FS:\setn \rightarrow \setv$ has the tree property iff
\[ \forall n\in\setn
\FS(n) \neq \empt \; \Longrightarrow \; [\FS(\parentf{n})] = [\vald] \]
wherever $\parent(n)$ is defined, that is, $n$ has a parent.
\end{mydef}

@@

An essential assumption in our model is that there is only
a \emph{single directory value} ($|\setd|=1$).
The realization that
this creates a symmetry between directories and empty nodes
and makes it possible to exploit a hidden symmetry in filesystem commands
that allows formulating a correct and complete reconciliation algorithm
is one of the main contributions of this paper.
In \cref{sec_multidir} we explore the applicability of our model given this assumption:
why applicability may not be affected and ways to relax this restriction.

@@

\myskip
In this paper
we fix the skeleton $\langle\setn,\parent\rangle$,
and the set of values $\setv$.
$\FS$ (with or without indices) denotes a filesystem,
and $n$, $m$ and $o$ are nodes in $\setn$.

>>>>

We model filesystems using functions with a set of potential filesystem paths or \emph{nodes} ($\setn$) as their domain,
and a set of possible contents or values ($\setv$) as their codomain.
$\setn$  serves as a namespace for the filesystem:
it contains all possible nodes, including the ones 
where the file system contains no file or directory,
and so for most filesystems it is infinite.
Nodes in $\setn$ are also arranged in trees by a parent function ($\parent$) as described in \cref{def_noderel}.

\begin{mydef}[Filesystems, $\FS$]
A filesystem $\FS$ is a function
mapping the set of nodes $\setn$ to values in $\setv$:
\[ \FS: \setn \rightarrow \setv. \]
\end{mydef}

@@

The tree-like structure of $\setn$ is determined by
an ancestor / descendant relation defined over the nodes
which arranges them in a disjoint union of rooted directed trees,
and which can be derived from the partial function $\parent$ yielding
the parent of a node in $\setn$ provided it exists.
Tao et al. \cite{TSR} describe a similar filesystem model, although
they also model inodes and restrict the filesystem to just a single tree.

\begin{mydef}[Ordering on $\setn$: $\parent$, $\descendant$, $\descendantEq$, $\unrel$]\mlabel{def_noderel}
The partial function $\parent:\setn\nrightarrow\setn$
returns the parent node of $n$,
and is undefined if $n$ is the root of a tree.

@@

We write $n\unrel m$, or $n$ and $m$ are \emph{incomparable},
iff $n\not\descendantEq m$ and $n\not\ancestorEq m$;
that is, incomparable nodes are on different branches or on different trees.
% none of the path names describing their location is an initial segment of the other.
\end{mydef}
By assumption that the $\parent$ function does not induce loops, and so
$\descendant$ is indeed a strict partial ordering.

@@

The combination of the set of nodes and the parent function,
$\langle \setn, \parent \rangle$, forms a \emph{skeleton}
which the filesystems populate with values from $\setv$.
As we require filesystem functions to be total,
we use a special value, $\empt\in\setv$, to indicate that the filesystem
is empty at a particular node, that is, there are no files or directories there.
We also assume that a filesystem has finitely many non-empty nodes,
and we consider any kind of metadata to be part of the values in $\setv$.

\begin{mydef}[Filesystem values: $\vald$, $\valfx$, $\empt$, $\eqclass{v}$]
The set of values $\setv$ is partitioned into directories, files, and the empty value $\empt$.
As usual, we write $[v]$ for the the equivalence class of $v\in\setv$ according to this partition,
which represents the type of the value.
File and directory values are denoted by $\valfx$ and $\vald$, respectively,
with or without indices, and so the partitioning is in fact $[\vald], [\valfx], [\empt]$.
\end{mydef}

@@

Every filesystem must have a so-called \emph{tree property}, which means that
if the filesystem is not empty at a node, and the node has a parent,
then there must be a directory at the parent node.
Using the notation introduced above, we can formally express this as follows.
\begin{mydef}[Tree property]
A filesystem $\FS$ has the tree property iff
\[ \forall n\in\setn:
\FS(n) \neq \empt \: \Longrightarrow \: \Big[\FS\big(\parentf{n}\big)\Big] = [\vald] \]
wherever $\parent(n)$ is defined, that is, $n$ has a parent.
\end{mydef}

@@

An essential additional assumption in our model is that there is only
a \emph{single directory value} ($|\setd|=1$).
The realization that
this creates a symmetry between directories and empty nodes
and makes it possible to exploit a hidden symmetry in filesystem commands
that allows formulating a correct and complete reconciliation algorithm
is one of the main contributions of this paper.
In \cref{sec_multidir} we explore the applicability of our model given this assumption:
why applicability may not be affected and ways to relax this restriction.

@@

In the rest of paper
we fix the skeleton $\langle\setn,\parent\rangle$,
and the set of values $\setv$.
$\FS$ (with or without indices) denotes a filesystem,
and $n$, $m$ and $o$ are nodes in $\setn$.

### 2018-07-12 p2/intro.tex
Style improvements

A central problem of command-based synchronizers 
during both update detection and reconciliation
is the ordering (scheduling) of commands.
If update detection is based on comparing the original state of the filesystem
to its new state, then we can easily collect a set of updates,
but we need to order these in a way that they could be applied
to the filesystem without causing an error 
(for example, a directory needs to be created before a file can be created under it).
Similarly, during reconciliation when we consider all the commands that
have been applied to the different replicas, we need to find a way of
merging these and creating a global ordering that we can apply
to the filesystems.
In fact, an insight in \cite{NREC} is that
if the ordering changes the effect of two commands,
then under certain circumstances they are not compatible and will give rise to conflicts.
Accordingly, in this paper command pairs, their commutativity and other properties
play a special role.

@@

The approach presented in this paper offers an improvement
over results described in \cite{NREC} in multiple ways.
We introduce a new set of commands that is symmetric and
captures more information about the updates,
and we restrict information content in directories
to exploit a further hidden symmetry in the command set.
These have the effect that reasoning about commands
becomes simpler as there are fewer edge cases,
and more powerful as 
predictions made by the reasoning are more accurate
due to the additional information content.
In fact, the new command set not only simplifies
our reconciliation algorithm, but also makes it maximal.
\cite{NREC} also lacked proofs that the update detection
and reconciliation algorithms it presented work as intended,
which we provide in the current paper.
While the results are intuitive, providing rigorous proofs
is far from trivial.
During the process, we define a number of useful concepts
and show their relationships and the properties they possess.
In our view these construct a special algebraic model
that is worthy of interest and of further research on its own.

The paper is organized as follows.
%
We start by defining a model of filesystems and a set of commands
that we will use to describe updates and modifications in \cref{sec_def}.
%
This is followed by investigating the properties and behaviors
of command pairs in \cref{section_axioms}.
%
\Cref{sec_update} describes update detection and ordering filesystem commands.
The main result of this section if that under some simple conditions,
a set of filesystem commands executed in any feasible order leads
to the same changes in the filesystem.
%
The reconciliation and conflict detection algorithm is defined in
\cref{sec_rec}, which merges two sets of commands.
We then proceed to prove that the output of reconciliation
is correct in the sense that 
it can be applied to the replicas without causing an error,
and it is maximal
inasmuch as no further updates can be applied under any circumstances.
In order to be able to do this, we introduce
domains of sets of command sequences,
and show that it has a number of highly convenient properties.
%
\Cref{sec_multidir,sec_algebra,sec_conclusion}
describe extending our model to widen its applicability;
outline directions for further work on the introduced algebraic system,
and provide the conclusion.
%
We discuss related work, comparing our results to other research in \cref{sec_relatedwork},
while
\cref{axiom_proof} contains technical proofs for propositions on the behavior of command pairs.

>>>>

A central problem of command-based synchronizers 
during both update detection and reconciliation
is the ordering (scheduling) of commands.
If update detection is based on comparing the original state of the filesystem
to its new state, then we can easily collect a set of updates,
but we need to order these in a way that they could be applied
to the filesystem without causing an error 
(for example, a directory needs to be created before a file can be created under it).
Similarly, during reconciliation when we consider all the commands that
have been applied to the different replicas, we need to find a way of
merging these and creating a global ordering that we can apply
to the filesystems.
In fact, an insight in \cite{NREC} is that
if the ordering changes the effect of two commands,
then under certain circumstances they are not compatible and will give rise to conflicts.
Accordingly, in this paper command pairs, their commutativity and other properties
play a significant role.

@@

The approach presented in this paper offers an improvement
over results described in \cite{NREC} in multiple ways.
We introduce a new set of commands that is symmetric and
captures more information about the updates,
and we restrict information content in directories
to exploit a further hidden symmetry in the command set.
These have the effect that reasoning about commands
becomes simpler as there are fewer edge cases,
and more powerful as 
predictions made by the reasoning are more accurate
due to the additional information content.
In fact, the new command set not only simplifies
our reconciliation algorithm, but also makes it maximal.
\cite{NREC} also lacked proofs that the update detection
and reconciliation algorithms it presented work as intended,
which we provide in the current paper.
While the results are intuitive, providing rigorous proofs
is far from trivial.
During the process, we define a number of auxiliary concepts
and show their relationships and the properties they possess.
In our view these construct a special algebraic model
that is worthy of interest and of further research on its own.

The paper is organized as follows.
%
We start by defining a model of filesystems and a set of commands
that we will use to describe updates and modifications in \cref{sec_def}.
%
This is followed by investigating the properties and behaviors
of command pairs in \cref{section_axioms}.
%
\Cref{sec_update} describes update detection and ordering filesystem commands.
The main result of this section is that under some simple conditions,
a set of filesystem commands executed in any feasible order leads
to the same changes in the filesystem.
%
The reconciliation and conflict detection algorithm is defined in
\cref{sec_rec}, which merges two sets of commands.
We then proceed to prove that the output of reconciliation
is correct in the sense that 
it can be applied to the replicas without causing an error,
and it is maximal
inasmuch as no further updates can be applied under any circumstances.
In order to be able to do this, we introduce
domains of sets of command sequences,
and show that it has a number of highly convenient properties.
%
\Cref{sec_multidir,sec_algebra,sec_conclusion}
describe extending our model to widen its applicability;
outline directions for further work on the introduced algebraic system,
and provide the conclusion.
%
We discuss related work, comparing our results to other research in \cref{sec_relatedwork},
while
\cref{axiom_proof} contains technical proofs for propositions on the behavior of command pairs.
